Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-10 paper/paper.tex
Fixing the section on implementation schemes

\subsection{A simple LGC Algorithm}
We shall call  a unit allocatable memory as a {\em  cell}.  A cell may
hold  a  basic value,  or  the constructor  \CONS\,  or  (parts of)  a
closure.  Apart from information to  identify the cell type, each cell
has the space  for two fields. The second of the  fields may be empty,
when, for example, the cell holds a basic value or a closure involving
a unary operator. \cred{Should we say more about the representation}.

\begin{figure*}[t!]
\label{fig:algos}
\hfill
\subfigure[]{\includegraphics[width=.27\linewidth]{lgc-scheme-1}}
\hfill
\subfigure[]{\includegraphics[width=.45\linewidth]{lgc-scheme-2}}
\hfill
\caption{(a) LGC with closure collection based on reachability  (b)
  LGC with liveness-based closure collection.}

@@

In  the simple algorithm  presented we  consult liveness  automata for
copying both \CONS cells and closures.  Copying \CONS cells are simple
and just  involve consulting  the liveness automaton  corresponding to
them  and  recursively  copying.  In  case of  closures,  it  is  more
complicated  as  they might  refer  to  data/closures  which might  be
created in  a different environment(function), all of  which might not
even exist  on the program  stack. To overcome  this we need  to carry
extra information with each closure, the program point at which it was
evaluated and  also the names of  the free variables  contained in the
closure.  This  gives us  precise  liveness  information  of the  free
variables of a closure but  involves run-time overhead of updating the
program  point correctly. To  avoid this  overhead we  go for  a safer
approximation by storing  the creation point of the  closure and using
it to check liveness.  This is safe since the liveness at the creation
point of the closure encompasses  all possible uses of the closure but
it might be very imprecise as at run-time the actual liveness required
might be way less than what is found at the creation point.

The advantage of  this method is that it collects  all dead cells even
if  they are  part  of a  closure  and hence  will  reduce the  memory
requirements and the number of garbage collections. The main drawbacks
of  this method  is  the fact  that  the algorithm  might do  multiple
traversals over  the same structure. Assume that  two references point
to the  same list, now using  reachability if it  is garbage collected
using  one reference  there is  no need  to traverse  using  the other
reference as everything  reachable from the root node  would have been
copied.  Unfortunately, this is not the case with liveness based GC as
the two references  might make different sections of  the list as live
and  copy  them.  Liveness  based  algorithm will  have  to  make  one
traversal  per reference.   Experimentally we  found that  this  was a
major source of overhead in the liveness based algorithm.  This can be
mitigated by  maintaining a  list of liveness  automata states  that a
particular heap  cell was traversed with  and ensuring that  we do not
repeat traversals  with the same  state. This requires that  each cell
maintain  a list  of state  which  potentially might  contain all  the
states.  Hence  this  is  not  a  feasible  solution.   \cred{repeated
  traversals if sharing}.

\subsection{Modified GC Algorithm}
Since the first algorithm requires a lot of bookkeeping and is not
precise, we have a second version of the algorithm
which does not use liveness for copying closures.
\begin{verbatim}
procedure copy (cell)
  IF{celltype is cons}
    IF cell is live &&
         cell is not copied using reachability
       new_cell <-  dup_heapcell(cell)
       new_car        <-  copy(cell.car)
       new_cell.car   <-  new_car
       new_cdr        <-  copy(cell.cdr)
       new_cell.cdr   <-  new_cdr
    ELSE
       cell <- NULL
    ENDIF
  ELSE
    IF {celltype is closure && cell is not copied}
       new_cell  <- copy_and_mark_reachable_cells(cell)
    ENDIF
  ENDIF
 copy_references_on_print_stack()
\end{verbatim}

In this  algorithm we consult  the liveness automata  only if it  is a
cons cell, for  closures we assume everything under it  to be live and
copy everything that is reachable. Although this might lead to copying
more  cells   it  is   faster  as  there   are  no   liveness  checks.
\comment{Once  experimental  results  are  obtained  compare  the  two
  algorithms here}  Mixing liveness based GC and  reachability has its
own set of pros and cons which we discuss further.\cred{hanging like a
  sore thumb}.

\subsubsection{Optimization for RGC for closures}
An advantage  of using  reachablity for closures  is that it  opens up
certain  opportunities  of  optimization.  As  we  saw  in  section[],
multiple traversals  over the same structure was  a major inefficiency
in the full liveness based  algorithm and overcoming it required a lot
of memory.  Using reachability for closures allows us to overcome this
partially using limited amount of  memory. Any heap cell which is part
of a closure  is copied using reachaibilty and  hence by maintaining a
single  flag in  each cell  to indicate  whether it  was  copied using
reachability    prevents   traversing    the   same    cell   multiple
times.  Experimental results  again indicate  that for  lazy languages
this gives a  substantial performance improvement.  \comment{Maybe add
  a diagram}

\subsubsection{Issues with mixing LGC and RGC}
A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closures or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
that any reference which is decalred  not live in a $GC_k$ will not be
derefernced during later execution.
% A reference copied using LGC could become part of a closure in a
later GC
% and copy wrong locations. Avoiding it by setting non-live references
to null after every GC.
% Proof for the correctness.
\subsection{Garbage collection for references on print stack}
Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function\comment{Refer to SPJ's
book}.
Printing of atomic values is simple and does not trigger a garbage
collection.
In case of \CONS\ cells, the \CAR\ part is first printed and then the \CDR\
part.
In a lazy language, both the \CAR\ and the \CDR\ parts could be closures,
thus
requiring evaluation. A garbage collection might be triggered while printing
the \CAR\ part or the \CDR\ part. Therefore, whenever a garbage collection is
triggered we have to consider any references that might be on the print stack
and copy them. Liveness automata does not exist for the print function as it
is not part of the executing program. Since the print function makes
everything
reachable from a cell live we can copy all cells reachable from the
reference.

Consider the case where a \CONS\ cells is being printed and a garbage
collection
is triggered when printing the \CDR\ part. In this case, since the refernce
to the
\CONS\ cells is still on the print stack both the \CAR\ and the \CDR\ parts
are
reachable but we know that the \CAR\ part has already been printed and can be
safely garbage collected. A liveness analysis of the print function would
have
given the same information. In our implementation we mimic the behavior of
the liveness analysis by using a flag in every \CONS\ cell which says whether
the \CAR\ part has been printed or not. For any \CONS\ cell if the \CAR\ part
has not been printed, a reference to the \CAR\ part will also be on the stack
but
the \CDR\ part has to be explicitly copied.
In case the \CAR\ part has been printed, then a reference to the \CDR\ part
will
be on the stack and since the \CAR\ part is not required anymore we do not
copy anything. This optimization subsumes tail call optimization. Again,
experimental results show that this optimization minimizes a lot of copying.
\cred{Very sketchy: Make another attempt to write}
\hrule
\comment{
Unlike eager languages where GC happens only at \CONS~ points or at
function calls, every \LET~ point is a potential GC point. The number
of
automata that needs to be stored is reduced by sharing automata along
all
the program points along an {\em E-path}.

By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.  An important difference in  GC of lazy language compared to
an  eager language is  the point  of garbage  collection.  In  case of
eager  languages only  \CONS~ points  and  function calls  are the  GC
points. In  a lazy  language every \LET~  statement is a  potential GC
point. Hence we need to store liveness automata corresponding to every
\LET~  statement in the  program. This  leads to  an explosion  in the
number of  automata. From our  observation that along an  {\em E-path}
all program points have the same liveness, we can reduce the number of
automata state required  by sharing the same automata  for all program
points along an {\em E-path}.} In  the example program, we only need to
create automata  for program  points \ldots, since  they are  the only
distinct {\em E-path}s in the program.

% Liveness for cons cells & reachability for closures

\comment{Mention the fact that liveness at a program point indicates
  what references will be used by the program eventually. The exact
  point where a reference will be used cannot be determined statically
due to the language being lazy. }
During GC, if the heap cell being processed is a cons cell, we consult
the  liveness automaton  corresponding to  it and  copy only  the live
paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
approximate the  liveness to be {\{0+1\}\*} and  copy everything under
the closure.
% Handling of mixture of LGC & RGC

% Difference between eager and lazy GC

% Doing GC for  references on the print stack - using  a flag to avoid
% copying non-live  cells Optimizing  GC for closures  - use  a copied
% using rgc flag for closures to avoid multiple traversals

@@

\subsection{Results}
\begin{figure}[t]
\centering
\begin{boxedminipage}{.75\textwidth}
{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\
\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
  then} {\bf skip}\\
\noindent \hspace*{.25cm}  {\bf else} {\bf if}
$\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR =
\mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\
\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$
{\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa},
\mathsf{dfa}(q,1),x.\CDR)$\\
\noindent \hspace*{.35cm}{\bf return} $y$
\end{boxedminipage}
\caption{Function       for       copying       a       root       set
  variable.}\label{algo:LGC-main-loop} \figrule

\end{figure}

@@

There are a couple of  issues of concern.  The garbage collection time
is larger in the case of LGC for some programs. The reason is that the
cost of consulting the liveness DFA may outweigh the combined benefits
of  fewer   garbage  collections   and  fewer  markings   per  garbage
collection.   The  other issue  is  illustrated  by  the program  {\tt
  lambda}    (memory   usage  shown   separately    in
Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
in Figure~\ref{fig:experimental-results}, the  number of
touched cells\footnote{These  are the  cells which are  visited during
  marking phase, often  more than once due to  sharing.} in this
example is much higher for
LGC.  This  increase is due to  excessive sharing among  heap nodes in
this program.  Note  that a node re-visited because  of sharing is not
explored  any   further  during  a  RGC   collection.   However,  this
curtailment cannot happen in  LGC because of the possibility that the
node,  re-visited in a  different liveness  state, may  mark a  set of
cells different from the earlier visit.
\begin{figure}[t]
\footnotesize

@@

\centerline{(b)}
\vskip -2mm
\caption{Experimental results comparing RGC and LGC. Table (a) gives
  data related to liveness analysis, and (b) gives garbage collection
  data. The average drag time is measured in allocated cons cells.}
\label{fig:experimental-results}
\normalsize
\end{figure}

\section{Collecting more garbage can never slow things down}
\label{sec:lgc-always-better}

%% Since garbage collection is  effectively asynchronous to the
allocator
%% thread, one might  worry as to how {\em  robust} our measurements
are.
%% For example,  while LGC would,  in general, collect more  garbage
than
%% RGC in the same heap state,  is it possible that on some programs,
LGC
%% might do  a larger number  of collections? We  prove a result  to
show
%% that   this  cannot   happen.    This  result   applies  to
classical
%% mark-and-sweep and copying garbage  collectors and, we hypothesize,
to
%% generational collectors too.

%% \begin{lemma}
%% For the  same mutator,  a liveness-based collector  can never  do
more
%% garbage collections than a reachability-based collector.
%% \end{lemma}

%% \begin{proof}Assume, as before,  that time is measured in terms of
the number of
%%   cons cells allocated. Let us run  two copies of the mutator with
the
%%   two    different   collectors    but    with   synchronous
memory
%%   allocations.   The   garbage    collections,   however,   happen
%%   asynchronously.

%%   To prove the lemma, it is enough to show the truth of the
following
%% statement: After
%%  every  LGC invocation,  the count of LGC invocations   is  no
greater
%% than  RGC invocations.   The  base case  holds  since the  first
%% invocations of both GCs happen at the same time.  Assume the
statement
%% to be true after $n$ invocations of LGC.  Since LGC copies a subset
of
%% reachable cells, its heap would contain no more cells than RGC heap
at
%% the end  of the $n$th invocation.   Thus either RGC  is invoked next
before
%% LGC, or LGC and RGC are both invoked next at the same time.  In
either
%% case, the statement holds after $n+1$ invocations of LGC.
%% \end{proof}

@@

Generated by bibtex from your ~.bib file.  Run latex,
then bibtex, then latex twice (to resolve references)
to create the ~.bbl file.  Insert that ~.bbl file into
the .tex source file and comment out
the command \texttt{{\char'134}thebibliography}.
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
\section{More Help for the Hardy}
The sig-alternate.cls file itself is chock-full of succinct
and helpful comments.  If you consider yourself a moderately
experienced to expert user of \LaTeX, you may find reading
it useful but please remember not to change it.
%\balancecolumns % GM June 2007
% That's all folks!

% {\color{red}
%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
% }

%% =======

%------------------------------------------------------------%
%\input{lazy-liveness}

>>>>

\subsection{A liveness-based garbage collection scheme}
We shall call  a unit allocatable memory as a {\em  cell}.  A cell can
hold  a  basic  value   ($\mathit{bas}$),  the  constructor  \CONS\  $(\mathit{cons~
arg_1~arg_2}$) or   a closure.  The closure, in  turn, can be  one of
($\mathit{unop~arg_1}$),  ($\mathit{ binop~arg_1~arg_2}$)  and  function  application
($\mathit{app~arg_1~arg_2})$.  Here each $\mathit{ arg_i}$ is a reference to another heap
cell or terminated through the function $\mathit{makenull}$.

\begin{figure*}[t!]
\label{fig:algos}
\hfill
\subfigure[]{\includegraphics[width=.30\linewidth]{lgc-scheme-1}}
\hfill
\subfigure[]{\includegraphics[width=.40\linewidth]{lgc-scheme-2}}
\hfill
\caption{(a) LGC with closure collection based on liveness.  (b)
  LGC with reachability-based closure collection.}

@@

Figure~\ref{fig:algos}  shows   a  liveness-based  garbage  collection
scheme.   The algorithm  assumes that  checking for  liveness involves
traversing the  liveness-DFAs stored as  a table and does  not mention
this explicitly.  Starting  with the root set, each  live reference is
explored   using  $\mathit{  lgcCopy}$.    Copying  \CONS\   cells  is
simple---it  just involves  copying  the cell  itself and  recursively
copying the \CAR\ and  the \CDR\ fields.  Liveness-based collection of
closures  is more  complicated since  our analysis  only  computes the
liveness of evaluated objects and not closures.

Let   us  consider   determining  the   liveness  of   a   closure  at
runtime. Assume  that the closure  $(app~f~y)$ has a  reference from
the root  set variable $x$. If $x$  is live,  then we  know that the
cell $(app~f~y)$ has  to be copied. However, to find out  the
liveness of $y$, could find out the program point which triggered the
evaluation of $(app~f~y)$ and pick up the liveness of $y$ from this
program point. However, this is only possible if the activation record
of the function which created this closure is still on the activation
stack. This does not always happen since a function could pack a
closure in a \CONS\ cell and return the \CONS\ cell.

An  alternative  could  be   to  carry  extra  information  with  each
closure---the {\em free variables in the closure and the program point
  at  which  it  was  evaluated}.   This  gives  us  precise  liveness
information  of the  free  variables  of a  closure  but involves  the
overhead of  updating the program point  information during execution.
To avoid this  overhead we go for a safe  approximation by storing the
{\em creation  point of the closure}  and using it  to check liveness.
This is safe  since the liveness at the creation  point of the closure
dominates all possible evaluation points  of the closure, but it might
be imprecise  as the actual  liveness might be  way less than  what is
found at the creation point.

Note that  when a reachability-based  collector visits a cell  for the
first time,  it is  marked.  Explorations starting  from the  cell are
curtailed if the cell is  visited subsequently due to sharing. This is
because the collector can assume  that the region of the heap starting
from  the  cell has  already  been  explored.  Since a  liveness-based
collector may  reach the cell each  time with a  different state, this
curtailment cannot  be done in  the case of  LGC. This is  an inherent
drawback  of a  liveness-based collector.   This can  be  mitigated by
maintaining a list of liveness  automata states that a particular heap
cell was traversed with and  ensuring that we do not repeat traversals
if we reach the cell in one of these states.

\subsection{Liveness-based collector with reachability-based closure
  collection}  Since the first  algorithm has  significant bookkeeping
overheads,  we have  a  second GC  scheme  that not  use liveness  for
copying closures.  In this  scheme, the liveness automata is consulted
only for \CONS\ cells; we assume everything under a closure to be live
and copy everything  that is reachable.

A consequence of copying everything under a closure is that we can end
up trying to copy things which are not live and corrupt the heap. This
can happen if a \CONS\  cell copied using liveness during a collection
subsequently  becomes   part  of  a  closure.    During  a  subsequent
collection,  the  collector  may  try  to copy  everything  under  the
\CONS\ cell thus copying heap  locations which might not be valid.  To
avoid  this, we  should ensure  that any  non-live references  are not
carried across  GC's.  We  do this  by scanning the  live part  of the
buffer after every  GC and setting to null any  reference which is not
live. This  is correct  because any reference  which is dead  during a
collection will not be dereferenced any further.

An advantage of  using reachablity for closures it  avoids the problem
of  multiple  explorations  from  the  same  cell.  This  is  done  by
maintaining  a flag in  each cell  to indicate  whether it  was copied
using  reachability.

\subsection{Garbage collection for references on print stack}
In lazy  languages the evaluation of  closures is driven  by the print
function~\cite{Jones87}.  Printing of  atomic values is
simple  and  does  not  trigger  a garbage  collection.   In  case  of
\CONS\ cells, the \CAR\ is first  printed followed by \CDR.  In a lazy
language, both \CAR\ and \CDR\ could be closures requiring evaluation,
and this  may trigger  a garbage collection.   When this  happens.  we
have to consider  any references that might be on  the print stack and
copy them. We extend the liveness analysis to the print function and
use its results during garbage collection.

%% Liveness automata  does not exist for  the print function  as it is
%% not part of the executing  program.  Since the print function makes
%% everything  reachable  from a  cell  live  we  can copy  all  cells
%% reachable from the reference.

%% Consider  the case  where a  \CONS\ cells  is being  printed  and a
%% garbage collection  is triggered when  printing the \CDR\  part. In
%% this case, since  the refernce to the \CONS\ cells  is still on the
%% print stack both the \CAR\ and the \CDR\ parts are reachable but we
%% know that the \CAR\ part has already been printed and can be safely
%% garbage collected. A liveness  analysis of the print function would
%% have given  the same information.   In our implementation  we mimic
%% the  behavior of the  liveness analysis  by using  a flag  in every
%% \CONS\ cell which  says whether the \CAR\ part  has been printed or
%% not. For any \CONS\ cell if  the \CAR\ part has not been printed, a
%% reference  to the  \CAR\ part  will also  be on  the stack  but the
%% \CDR\ part has to be explicitly copied.  In case the \CAR\ part has
%% been printed,  then a reference  to the \CDR\  part will be  on the
%% stack and  since the \CAR\ part  is not required anymore  we do not
%% copy anything.  This  optimization subsumes tail call optimization.
%% Again, experimental results show that this optimization minimizes a
%% lot of copying.  \cred{Very sketchy: Make another attempt to write}
%% \hrule

%% \comment{ Unlike  eager languages where  GC happens only  at \CONS~
%%   points or at function calls,  every \LET~ point is a potential GC
%%   point.  The number of automata that needs to be stored is reduced
%%   by sharing  automata along all  the program points along  an {\em
%%   E-path}.

%% By the time the program starts executing, the liveness automata are
%% created  and stored  in a  file.  This  is re-used  every  time the
%% program executes.   An important difference in GC  of lazy language
%% compared to an  eager language is the point  of garbage collection.
%% In case  of eager languages  only \CONS~ points and  function calls
%% are the  GC points. In a  lazy language every \LET~  statement is a
%% potential  GC  point. Hence  we  need  to  store liveness  automata
%% corresponding to  every \LET~ statement in the  program. This leads
%% to an  explosion in  the number of  automata. From  our observation
%% that  along  an {\em  E-path}  all  program  points have  the  same
%% liveness, we  can reduce the  number of automata state  required by
%% sharing  the same  automata for  all program  points along  an {\em
%% E-path}.} In the  example program, we only need  to create automata
%% for program  points \ldots, since  they are the only  distinct {\em
%% E-path}s in the program.

@@

\subsection{Results}
%% \begin{figure*}[t]
%% \centering
%% \begin{boxedminipage}{.75\textwidth}
%% {\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\
%% \noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
%% \noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
%%   then} {\bf skip}\\
%% \noindent \hspace*{.25cm}  {\bf else} {\bf if}
%% $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR =
%% \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\
%% \noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$
%% {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa},
%% \mathsf{dfa}(q,1),x.\CDR)$\\
%% \noindent \hspace*{.35cm}{\bf return} $y$
%% \end{boxedminipage}
%% \caption{Function       for       copying       a       root       set
%%   variable.}\label{algo:LGC-main-loop} \figrule

%% \end{figure*}

@@

There are a couple of  issues of concern.  The garbage collection time
is larger in the case of LGC for some programs. The reason is that the
cost of consulting the liveness DFA may outweigh the combined benefits
of  fewer   garbage  collections   and  fewer  markings   per  garbage
collection.   The  other issue  is  illustrated  by  the program  {\tt
  lambda}    (memory   usage  shown   separately    in
Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
in Figure~\ref{fig:experimental-results}, the  number of
touched cells\footnote{These  are the  cells which are  visited during
  marking phase, often  more than once due to  sharing.} in this
example is much higher for
LGC.  This  increase is due to  excessive sharing among  heap nodes in
this program.  Note  that a node re-visited because  of sharing is not
explored  any   further  during  a  RGC   collection.   However,  this
curtailment cannot happen in  LGC because of the possibility that the
node,  re-visited in a  different liveness  state, may  mark a  set of
cells different from the earlier visit.
\begin{figure*}[t]
\footnotesize

@@

\centerline{(b)}
\vskip -2mm
\caption{Experimental results comparing RGC and LGC. Table (a) gives
  data related to liveness analysis, and (b) gives garbage collection
  data. The average drag time is measured in allocated cons cells.}
\label{fig:experimental-results}
\normalsize
\end{figure*}

%% \section{Collecting more garbage can never slow things down}
%% \label{sec:lgc-always-better}

%% Since garbage collection is  effectively asynchronous to the
%% allocator
%% thread, one might  worry as to how {\em  robust} our measurements
%% are.
%% For example,  while LGC would,  in general, collect more  garbage
%% than
%% RGC in the same heap state,  is it possible that on some programs,
%% LGC
%% might do  a larger number  of collections? We  prove a result  to
%% show
%% that   this  cannot   happen.    This  result   applies  to
%% classical
%% mark-and-sweep and copying garbage  collectors and, we hypothesize,
%% to
%% generational collectors too.

%% \begin{lemma}
%% For the  same mutator,  a liveness-based collector  can never  do
%% more
%% garbage collections than a reachability-based collector.
%% \end{lemma}

%% \begin{proof}Assume, as before,  that time is measured in terms of
%% the number of
%%   cons cells allocated. Let us run  two copies of the mutator with
%% the
%%   two    different   collectors    but    with   synchronous
%% memory
%%   allocations.   The   garbage    collections,   however,   happen
%%   asynchronously.

%%   To prove the lemma, it is enough to show the truth of the
%% following
%% statement: After
%%  every  LGC invocation,  the count of LGC invocations   is  no
%% greater
%% than  RGC invocations.   The  base case  holds  since the  first
%% invocations of both GCs happen at the same time.  Assume the
%% statement
%% to be true after $n$ invocations of LGC.  Since LGC copies a subset
%% of
%% reachable cells, its heap would contain no more cells than RGC heap
%% at
%% the end  of the $n$th invocation.   Thus either RGC  is invoked next
%% before
%% LGC, or LGC and RGC are both invoked next at the same time.  In
%% either
%% case, the statement holds after $n+1$ invocations of LGC.
%% \end{proof}

@@

%% Generated by bibtex from your ~.bib file.  Run latex,
%% then bibtex, then latex twice (to resolve references)
%% to create the ~.bbl file.  Insert that ~.bbl file into
%% the .tex source file and comment out
%% the command \texttt{{\char'134}thebibliography}.
%% % This next section command marks the start of
%% % Appendix B, and does not continue the present hierarchy
%% \section{More Help for the Hardy}
%% The sig-alternate.cls file itself is chock-full of succinct
%% and helpful comments.  If you consider yourself a moderately
%% experienced to expert user of \LaTeX, you may find reading
%% it useful but please remember not to change it.
%% %\balancecolumns % GM June 2007
%% % That's all folks!

%% % {\color{red}
%% %   (Note 3.\ includes $e_\mainpgm$ treated as the body of
%% %   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
%% % }

%% %% =======

%% %------------------------------------------------------------%
%% %\input{lazy-liveness}
